---
title: "Untitled"
author: "Boyko Amarov"
date: "3/14/2022"
output: html_document
---

Ctrl-Alt-i

```{r}
1 + 4
3 - 2
2 * 8
2 / 8
2 ^ 4
```


```{r}
## Shortcut for the assignment operator: Alt - 
y <- 34
x <- c(1, 3)
x
```

```{r}
## Average of two number
mean(y)
```


```{r}
## Opens the documentation for a function
?mean
```

## String values

```{r}
z <- "Hello, world!"
# z + 1
```


## Logical values

```{r}
xTrue <- 2 < 3
xFalse <- 2 == 3

TRUE + FALSE
```

## Data

```{r}
library(tidyverse)
```


```{r}
earnings <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```

- _height_ (numeric): height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1 if a person is male, 0 otherwise
- _earn_ (numeric): annual income in USD
- _earnk_ (numeric): annual income in 1000 USD
- _ethnicity_ (character): ethnicity
- _age_ (numeric): Age of the person in years


```{r}
##
earnings <- earnings %>%
  mutate(
    height_cm = 2.54 * height,
    weight_kg = 0.45 * weight
  )
```

## Basic data summaries

### Location

```{r}
## Average height (in cm)
mean(earnings$height_cm)
```


$$
x = (x_1, x_2,\ldots, x_n)\\
\bar{x} = \frac{1}{n}\sum_{i = 1}^{n} x_i
$$
$n = 1816$.

### Variation

```{r}
max(earnings$height_cm)
min(earnings$height_cm)
```
```{r}
max(earnings$height_cm) - min(earnings$height_cm)
```


```{r}
range(earnings$height_cm)
```

```{r}
summary(earnings$height_cm)
```

Variance

$$
S^2_x = \frac{1}{n - 1} \sum_{i = 1}^{n}(x_i - \bar{x})^2\\
S_x = \sqrt{S^2_x}: \text{Standard deviation}
$$

```{r}
var(earnings$height_cm)
```

```{r}
sqrt(var(earnings$height_cm))
```

```{r}
sd(earnings$height_cm)
```

