---
title: "Седмица 2"
author: "Boyko Amarov"
date: "3/8/2022"
output: html_document
---

```{r}
1 + 1
2 * 3
2 - 5
2 ^ 3
2 / 4
```

```{r}
# Shortcut: Alt -
x <- 2
x + 4
z <- 7
```

```{r}
y <- "Добър ден, свят!"
z
```

```{r}
TRUE
FALSE
```

```{r}
2 > 3
```

За вмъкване на блок от код в текста: Ctrl Alt i

```{r}
xTrue <- TRUE
xFalse <- FALSE
xTrueString <- "TRUE"
```

```{r}
xTrue + 1
xFalse + 1
```

```{r}
xTrueString + 1
```

# Data frames

```{r}
# install.packages("tidyverse")
```

# Data

- _height_ (numeric): height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1 if a person is male, 0 otherwise
- _earn_ (numeric): annual income in USD
- _earnk_ (numeric): annual income in 1000 USD
- _ethnicity_ (character): ethnicity
- _age_ (numeric): Age of the person in years


```{r}
library(tidyverse)

earnings <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```

## Data mutations

New column called "height_cm"

```{r}
earnings <- earnings %>%
  mutate(
    height_cm =  2.54 * height,
    weight_kg = 0.45 * weight
  ) %>%
  select(height_cm, weight_kg, ethnicity)
```

## Data summaries

```{r}
mean(earnings$height_cm)
```


$$
x_1, x_2,\ldots, x_n\\
\bar{x} = \frac{1}{n}\sum_{i = 1}^{n} x_i
$$







