---
title: 'Week2: basic data summaries'
author: "Boyko Amarov"
date: "3/8/2021"
output: html_document
---

## Data import

```{r, message=FALSE}
## Load packages
library(tidyverse)

## Read the data
earnings <- read_csv(
  "https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv"
  ) %>%
  mutate(
    gender = ifelse(male == 1, "Мänner", "Frauen")
  )
```

Data description

- _height_ (numeric): height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1 if a person is male, 0 otherwise
- _earn_ (numeric): annual income in USD
- _earnk_ (numeric): annual income in 1000 USD
- _ethnicity_ (character): ethnicity
- _age_ (numeric): Age of the person in years

Create a new column holding the height in cm. We use the `mutate` function for this task.

```{r}
##
## earnings <- mutate(earnings, height_cm = 2.54 * height)
## Equivalently using pipes:
earnings <- earnings %>%
  mutate(
    height_cm = 2.54 * height,
    weight_kg = 0.45 * weight
  )
```

First look into the data set.

```{r}
glimpse(earnings)
```

Arithmetic average

```{r}
mean(earnings$height_cm)
```

The average height of the persons was 169 cm.

```{r}
summary(earnings$height_cm)
```

Der Median is 167.6 cm: ungefähr die Hälfte der Kunden haben eine Größe, die
kleiner ist als 167.6 cm.

1. Quartil ist 162.6 cm: ungefähr 1/4 der Kunden haben eine Größe, die
kleiner ist als 162.6 cm.
3. Quartil ist 175.9 cm: ungefähr 3/4 der Kunden haben eine Größe, die
kleiner ist als 175.9 cm.


## Box-Whiskers plot

```{r}
## ggplot: Grammar of Graphics

ggplot(
  ## Data source
  data = earnings,
  ## Aesthetics
  aes(
    ## x-axis
    x = height_cm
  )
) + 
  geom_boxplot() +
  labs(
    title = "Boxplot",
    x = "Größe (cm)"
  )
```

```{r}
ggplot(
  data = earnings, 
  aes(
    x = height_cm, 
    y = gender
    )
  ) +
    geom_boxplot() +
    labs(
      x = "Height (cm)",
      y = "Gender"
    )
```


```{r}
ggplot(data = earnings, aes(x = height_cm, y = ethnicity)) + geom_boxplot()
```


```{r}
## Extreme case of no variation in the data (all values equal)
summary(c(2, 2, 2, 2, 2, 2, 2))
ggplot(data = data.frame(x = c(2, 2, 2, 2, 2, 2, 2)), aes(x = x)) + geom_boxplot()
```

1. Location: (measured by the median)
2. Spread: (distance between min/max and between the 1st and the 3the quartiles)
3. Outliers: observations that are different from the rest of the observations (persons)


```{r}
earnings1 <- earnings %>%
  mutate(
    male1 = male + 1
  ) %>%
  # select(gender)
  select(male, male1, gender, ethnicity, height_cm)
```


```{r}
summary(earnings1$male)
```


```{r}
summary(earnings1$male1)
```


For _categorical_ variables we use frequency tables as a summary.

```{r}
table(earnings$gender)
```


```{r}
table(earnings$ethnicity)
```

```{r}
ggplot(data = earnings, aes(x = ethnicity)) + geom_bar()
```

There are 180 black persons in the dataset.

## Measurement scales

Categorical scales:

1. **Nominal**: Values of the variable have no numeric meaning. Values are only labels indicating group membership.
2. **Ordinal**: Values of the variable are just labels (as is the case with the nominal scale), but the labels are ordered.
 school grades
 "2" "3" "4" "5" "6"

Numeric scales
<!-- 3. **Interval**: -->
4. **Ratio scale**: Numeric scale, ratios between values make sense.


## Zusammenhang zwischen zwei Variablen

Streudiagramm

```{r}
mean(earnings$weight_kg, na.rm = TRUE)
```

```{r}
NA + 2
```

```{r}
## Compute the mean by simply omitting the missing values
mean(earnings$weight_kg, na.rm = TRUE)
mean(earnings$height_cm)
## NA: missing/undefined value
```

```{r}
ggplot(
  data = earnings,
  aes(
    x = weight_kg, 
    y = height_cm
    )
  ) +
  geom_point(
    position = "jitter",
    size = 1 / 2,
    alpha = 0.5
  ) +
  labs(
    title = "Scatterplot",
    y = "Height (cm)",
    x = "Weight (kg)"
  ) +
  geom_smooth(method = "lm")
```
