---
title: "Simple Linear Model"
author: "Boyko Amarov"
date: "3/28/2022"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
invoices <- read.delim('https://raw.githubusercontent.com/feb-uni-sofia/econometrics2020-solutions/master/data/invoices.txt')
```


```{r}
glimpse(invoices)
```

Variables description:

- `Day` (numeric): day
- `Invoices` (numeric): number of invoices
- `Time` (numeric): Time needed to process the invoices (hours)

Task is to predict the time needed to process 50, 120, 200, 250, 400 invoices.

```{r}
summary(invoices$Time)
```


```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_vline(xintercept = c(50, 120, 201, 250, 400), lty = 2, alpha = 0.5) +
  scale_x_continuous(breaks = c(50, 120, 201, 250, 400)) +
  geom_abline(slope = 0.015, intercept = 0.1) +
  geom_abline(slope = 0.001, intercept = 0.1, color = "firebrick") +
  ylim(c(0, 7))

# +
  # geom_vline(xintercept = 120, lty = 2) +
  # geom_hline(yintercept = 2.110) +
  # geom_abline(intercept = 0.1, slope = 0.015, colour = "steelblue2")
```

Estimate the expected processing time for 50, 120, 201, 250, 400 invoices (3 estimates).

Strategy 1: Use the average processing time for the predictions

$$
i = 1,\ldots,n = 30\\
x_i: \text{ number of invoices on day } i \\
y_i: \text{ processing time on day } i \\
$$
Assume a simple (linear) functional relationship

$$
y_i = 0.1 + 0.015x_i
$$

This equation cannot describe the relation between the processing time and the
number of invoices is too restrictive. It implies that all dots in the scatterplot
lie on the line with intercept 0.1 and with a slope equal to 0.015.


$$
y_i = 0.1 + 0.015x_i + u_i
$$

The extra term $u_i$ accounts for the deviations of the observed processing times from the line.

Assume that $u_i$ follows a normal distribution with expected value (mean) of zero and variance $\sigma^2$. We also that $u_i$ are independent and that $u_i$ are independent of $x_i$.

$$
u_i \sim N(0, \sigma^2)
$$

Under the assumption that $E(u_i) = 0$ we can compute the conditional expected
value of $y_i$. We use the linearity of the expected value:

$$
E(y_i) = E(0.1 + 0.015x_i + u_i)\\
E(y_i) = E(0.1) + E(0.015x_i) + E(u_i) \\
E(y_i) = 0.1 + 0.015x_i + 0 \\
E(y_i) = 0.1 + 0.015x_i
$$
The _expected_ processing time $E(y_i|x_i)$ is equal to $0.1 + 0.015x_i$.

$$
y_i = \underbrace{0.1 + 0.015x_i}_{\text{Systematic part}} + \underbrace{u_i}_{\text{Random part}}
$$
## Simulation

Normal distribution

```{r}
# r: random, norm: normal
# mean: expected value of the distribution, 
# sd: standard deviation of the distribution = square root of the variance
u <- rnorm(1000, mean = 0, sd = 1)
u
```

```{r}
tibble(u) %>%
  ggplot(aes(x = u)) +
    geom_histogram(bins = 20)
```


```{r}
n <- 30
x <- round(seq(10, 250, length.out = n), 0)
## Select values at random from a standard 
## normal distribution, i.e. mean = 0, standard dev. = 0.5
u <- rnorm(n, mean = 0, sd = 0.5)
y <- 0.1 + 0.015 * x + u

tibble(x = x, y = y) %>%
  ggplot(aes(x = x, y = y)) + 
    geom_point() +
    geom_abline(intercept = 0.1, slope = 0.015) +
    xlim(c(0, 250))
```





$$
y[\text{hours}] = 0.1[\text{hours}] + \underbrace{0.015\left[\frac{\text{hours}}{\text{invoice}}\right] x[\text{numb. invoices}]}_{\text{hours}}\\
x = 0 \implies y [\text{hours}] = 0.1 [\text{hours}]
$$


$$
50 [m^2] \text{ room}\\
\text{rent } 2[\frac{BGN}{m^2}]\\
2 [\frac{BGN}{m^2}] \times 50 [m^2] = 100 [BGN]
$$

$$
y[\text{costs}] = 0.1[\text{fixed costs}] + 0.015[\text{marginal cost}]x[\text{production}]
$$

$$
y_{x = 120} = 0.1 + 0.015 \times 120 = 1.9 [\text{hours}]
$$

$$
y = \underbrace{0.1}_{\text{intercept}} + \underbrace{0.015}_{\text{slope}}x
$$

$$
y_i  = \underbrace{0.1 + 0.015x_i}_{\text{line, systematic part}} + \underbrace{e_i}_{\text{deviation from the line, stochastic (random) part}}
$$



```{r}
x <- 190 ## Number of invoices
e <- runif(1, min = -1, max = 1) ## Uniform distribution on [-1, 1]
e
y <- 0.1 + 0.015 * x + e
y
```

```{r}
rand_numbers <- runif(10, min = -1, max = 1)
# rand_numbers
paste("Average result from a game:", mean(rand_numbers))
rand_numbers
```

Assume that the deviation part is zero on average.

$$
E(e_i) = 0 \text{ assumption} \implies \\
E(y_i) = E(0.1 + 0.015x_i + e_i) =\\ 
E(y_i) = E(0.1) + E(0.015x_i) + E(e_i) = \\
E(y_i) = 0.1 + 0.015x_i
$$

## Coefficient estimation

$$
y = 0.1 + 0.015x + e_i
$$
Predicted y (expected value of y) for every observation

$$
\hat{y}_i = 0.1 + 0.015x_i
$$


```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_abline(intercept = 0.6, slope = 0.015, colour = "steelblue2") +
  geom_abline(intercept = 0.3, slope = 0.02, colour = "steelblue2", alpha = 0.5) +
  geom_abline(intercept = 0.6, slope = 0.010, colour = "steelblue2", alpha = 0.5) +
  geom_abline(intercept = 2, slope = 0.015, colour = "firebrick")
```


$$
y_i - \hat{y_i} \text{ residual for day } i
$$
```{r}
## Illustration only
invoices <- invoices %>%
  mutate(
    y_predicted_1 = 0.1 + 0.015 * Invoices
  )
ggplot(data = invoices, aes(x = Invoices, y = Time)) +
  geom_point(size = 2) +
  geom_segment(
    aes(xend = Invoices, yend = y_predicted_1),
    size = 1 / 4
  ) +
  # geom_point(data = data.frame(x = c(201, 201), y = c(4.5, 2)), aes(x = x, y = y), shape = 1) +
  # geom_segment(data = data.frame(x = c(201, 201), y = c(4.5, 2)),
  #              aes(x = x, y = y, xend = 201, yend = 0.1 + 0.015 * 201), lty=2) +
  geom_abline(intercept = 0.1, slope = 0.015, color = "steelblue")
```
How to aggregate each individual deviation (residual)? One possible approach is to simply sum the deviations

$$
\sum_{i = 1}^{n} (y_i - \hat{y}_i)^2
$$


$$
y_i = \alpha + \beta x_i + e_i\\
E(y_i) = \alpha + \beta x_i
$$


Find guesses for $\alpha$ and $\beta$ based on the data, so that the distance between
the predictions and the observations is as small as possible. Let's call our
"guesses" for $\alpha$ $\hat{\alpha}$ and for $\beta$ $\hat{\beta}$

$$
\hat{y}_i = \hat{E}(y_i) = \hat{\alpha} + \hat{\beta}x_i \text{ guessed expected value}\\
(\hat{\alpha}^{OLS}, \hat{\beta}^{OLS}) = \min_{\hat{\alpha}, \hat{\beta}} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2
$$

The result of this minimisation is what we call the OLS (ordinary least squares) estimator for $\alpha$ and for $\beta$.

## Application in R

$$
\underbrace{y_i}_{Time} = \alpha + \beta \underbrace{x_i}_{Invoices} + e_i
$$

```{r}
## lm: linear model
lm(Time ~ 1 + Invoices, data = invoices)
```

Estimated equation using OLS.

$$
\hat{\alpha} = 0.64\\
\hat{\beta} = 0.011\\
\hat{y}_i =  0.64 + 0.011 x_i
$$
We estimate the expected processing time at $x_i$ invoice to be the sum the estimated fixed costs (0.64 hours) plus the estimated marginal costs (0.011 hours/invoice) times the number of invoices ($x_i$).

The expected processing time increases by 0.011 *hours* for each additional invoice.

The expected processing time at zero invoices is 0.64 _hours_.

## Plot the OLS estimated line

```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_abline(intercept = 0.1, slope = 0.015, colour = "steelblue2") + 
  geom_abline(intercept = 0.64, slope = 0.011, colour = "firebrick")
```

```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_smooth(method = "lm")
```