---
title: "Simple Linear Model"
author: "Boyko Amarov"
date: "3/28/2022"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
invoices <- read.delim('https://raw.githubusercontent.com/feb-uni-sofia/econometrics2020-solutions/master/data/invoices.txt')
```


```{r}
glimpse(invoices)
```

Variables description:

- `Day` (numeric): day
- `Invoices` (numeric): number of invoices
- `Time` (numeric): Time needed to process the invoices (hours)

Task is to predict the time needed to process 50, 120, 201, 250, 400 invoices.


```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_vline(xintercept = c(50, 120, 201, 250, 400), lty = 2, alpha = 0.5) +
  scale_x_continuous(breaks = c(50, 120, 201, 250, 400)) +
  geom_abline(slope = 0.015, intercept = 0.1) +
  ylim(c(0, 7))
# +
  # geom_abline(slope = 0.001, intercept = 0.1, color = "firebrick") +


# +
  # geom_vline(xintercept = 120, lty = 2) +
  # geom_hline(yintercept = 2.110) +
  # geom_abline(intercept = 0.1, slope = 0.015, colour = "steelblue2")
```

Estimate the expected processing time for 50, 120, 201, 250, 400 invoices (3 estimates).

Strategy 1: Use the average processing time for the predictions

$$
i = 1,\ldots,n = 30\\
x_i: \text{ number of invoices on day } i \\
y_i: \text{ processing time on day } i \\
$$
Assume a simple (linear) functional relationship

$$
y_i = 0.1 + 0.015x_i
$$

This equation cannot describe the relation between the processing time and the
number of invoices is too restrictive. It implies that all dots in the scatterplot
lie on the line with intercept 0.1 and with a slope equal to 0.015.


$$
y_i = 0.1 + 0.015x_i + u_i
$$

The extra term $u_i$ accounts for the deviations of the observed processing times from the line.

Assume that $u_i$ follows a normal distribution with expected value (mean) of zero and variance $\sigma^2$. We also that $u_i$ are independent and that $u_i$ are independent of $x_i$.

$$
u_i \sim N(0, \sigma^2)
$$

Under the assumption that $E(u_i) = 0$ we can compute the conditional expected
value of $y_i$. We use the linearity of the expected value:

$$
E(y_i) = E(0.1 + 0.015x_i + u_i)\\
E(y_i) = E(0.1) + E(0.015x_i) + E(u_i) \\
E(y_i) = 0.1 + 0.015x_i + 0 \\
E(y_i) = 0.1 + 0.015x_i
$$
The _expected_ processing time $E(y_i|x_i)$ is equal to $0.1 + 0.015x_i$.

$$
y_i = \underbrace{0.1 + 0.015x_i}_{\text{Systematic part}} + \underbrace{u_i}_{\text{Random part}}
$$
$$
y_i = E(y_i) + u_i
$$

## Simulation

Normal distribution

```{r}
# r: random, norm: normal
# mean: expected value of the distribution, 
# sd: standard deviation of the distribution = square root of the variance

u <- rnorm(1000, mean = 0, sd = 1)
# u
```

```{r}
mean(u)
```


```{r}
tibble(u) %>%
  ggplot(aes(x = u)) +
    geom_histogram(bins = 20)
```


```{r}
n <- 30
x <- round(seq(10, 250, length.out = n), 0)
## Select values at random from a standard 
## normal distribution, i.e. mean = 0, standard dev. = 0.5
u <- rnorm(n, mean = 0, sd = 0.5)
y <- 0.1 + 0.015 * x + u

tibble(x = x, y = y) %>%
  ggplot(aes(x = x, y = y)) + 
    geom_point() +
    geom_abline(intercept = 0.1, slope = 0.015) +
    xlim(c(0, 250))
```
